Anomaly detection with kernel density estimation on manifolds

### 2022-02-26 paper/paper.Rmd
100d simulation process writing

In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as $d=2$, same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=10,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.

## 2-D Meta data from a Gaussian Mixture Model {#twodgaussian}

@@

## 5-D data embedding in 100-D space {#fivedgaussian}

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space. First, we simulate $N=10,000$ points from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$. In order to manually generate anomalies, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$.
The fifth dimension is calculated to satisfy $x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2$ where $r$ is set as $7$. Similarly, the Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package *tourr* [REFERENCE].
Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size $N$ (denote the transpose of the data matrix as $X$) to get rid of the zeros so that it could be passed to the manifold learning algorithms. 
The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $A$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
For any real matrix $A$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $Q$ and $R$ so that $A = QR$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $Q^\prime Q = I$, and $R$ is an upper triangular matrix. Matrix $Q$ is used as the rotation matrix and it satisfies 
$X^\prime X=(QX)^\prime(QX)$, where the rotated data matrix $(QX)^\prime$ is now the input for the manifold learning algorithms.

```{r fivedmeta, fig.align = 'center', out.width = "80%", fig.cap="Scatterplot display of the animation of a 5-D tour path.", echo = FALSE}
knitr::include_graphics("figures/tourr_5d_semisphere.png")
```

>>>>

In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as $d=2$, same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.

## 3-D mapping from a 2-D Gaussian Mixture Model {#twodgaussian}

@@

## 100-D mapping from a 5-D semi-hypersphere {#fivedgaussian}

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in $d=5$ with manifold learning. First, we simulate $N=2,000$ points, $(\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime$, from a 4-dimensional Gaussian mixture model with two mixture components, $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$, where $\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime$, $\pmb{\Sigma}_1 = diag(1,1,1,1)$, and $\pmb{\Sigma}_1 = diag(2,2,2,2)$. In order to manually add anomalies to be distant points from the means, the mixture proportions are set as $\pi_1=0.99$ and $\pi_2=0.01$. 
The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation $X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2$ where $X_5>0$ and $r$ is set as $7$. 
The Gaussian mixture densities could be calculated using Equation \@ref(eq:gmm) as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package *tourr* [REFERENCE]. The round and triangular point shapes index the two mixture components $\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)$ and $\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)$, and the colors shows the distance between the simulated $4-D$ data point from Gaussian mixture model and the kernel means $(0, 0, 0, 0)^\prime$. The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}. It can be seen that the most distant points are in triangular shape, meaning that they are simulated from $\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)$. The dark colors also indicate that they are the true outliers because of their low densities.

Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the $100$-dimensional data of size $N$ (denote the transpose of the data matrix as $\pmb{X}_0$ with dimension $100 \times N$) to get rid of the zeros so that it could be passed to the manifold learning algorithms. 
The rotation matrix is derived from the QR decomposition of a $100\times 100$ matrix $\pmb{A}$ with all components randomly generated from a uniform distribution $\mathcal{U}(0,1)$. 
For any real matrix $\pmb{A}$ of dimension $p\times q$, the QR decomposition could decompose the matrix into the multiplication of two matrix $\pmb{Q}$ and $\pmb{R}$ so that $\pmb{A} = \pmb{QR}$, where the dimension of $Q$ is a matrix with unit norm orthogonal vectors, $\pmb{Q}^\prime \pmb{Q} = \pmb{I}$, and $\pmb{R}$ is an upper triangular matrix. Matrix $\pmb{Q}$ satisfies 
$\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)$, meaning that the distances between data points in $\pmb{X}_0^\prime$ is equivalent to that of $(\pmb{QX}_0)^\prime$. Therefore, we use matrix $Q$ as the rotation matrix for  where the rotated data matrix $\pmb{X} = (\pmb{QX}_0)^\prime$ of dimension $N \times 100$ is now the input data for the manifold learning algorithms. Again, we set the embedding dimension to be equal to the meta data dimension $d=5$.

```{r fivedmeta, fig.align = 'center', out.width = "70%", fig.cap="Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.", echo = FALSE}
knitr::include_graphics("figures/tourr_5d_semisphere.png")
```

### 2022-02-26 paper/paper.tex
100d simulation process writing

In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=10,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

\hypertarget{twodgaussian}{%
\subsection{2-D Meta data from a Gaussian Mixture Model}\label{twodgaussian}}

@@

\hypertarget{fivedgaussian}{%
\subsection{5-D data embedding in 100-D space}\label{fivedgaussian}}

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere and transform it into 100-dimensional space. First, we simulate \(N=10,000\) points from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\). In order to manually generate anomalies, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy \(x_1^2 + x_2^2 + x_3^2 + x_4^2 + x_5^2 = r^2\) where \(r\) is set as \(7\). Similarly, the Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}.
Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the 100-dimensional data of size \(N\) (denote the transpose of the data matrix as \(X\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(A\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
For any real matrix \(A\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(Q\) and \(R\) so that \(A = QR\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(Q^\prime Q = I\), and \(R\) is an upper triangular matrix. Matrix \(Q\) is used as the rotation matrix and it satisfies
\(X^\prime X=(QX)^\prime(QX)\), where the rotated data matrix \((QX)^\prime\) is now the input for the manifold learning algorithms.

@@

{\centering \includegraphics[width=0.8\linewidth]{figures/tourr_5d_semisphere}

@@

\caption{Scatterplot display of the animation of a 5-D tour path.}\label{fig:fivedmeta}
\end{figure}

>>>>

In this section, we exam two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purpose, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), same as the meta data dimension. This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

\hypertarget{twodgaussian}{%
\subsection{3-D mapping from a 2-D Gaussian Mixture Model}\label{twodgaussian}}

@@

\hypertarget{fivedgaussian}{%
\subsection{100-D mapping from a 5-D semi-hypersphere}\label{fivedgaussian}}

As a high-dimensional experiment, we generate the meta data from a 5-dimensional semi-hypersphere, transform it into a 100-dimensional space, and then embed it in \(d=5\) with manifold learning. First, we simulate \(N=2,000\) points, \((\pmb{X}_1, \pmb{X}_2, \pmb{X}_3, \pmb{X}_4)^\prime\), from a 4-dimensional Gaussian mixture model with two mixture components, \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\), where \(\pmb{\mu}_1 = \pmb{\mu}_2 =(0, 0, 0, 0)^\prime\), \(\pmb{\Sigma}_1 = diag(1,1,1,1)\), and \(\pmb{\Sigma}_1 = diag(2,2,2,2)\). In order to manually add anomalies to be distant points from the means, the mixture proportions are set as \(\pi_1=0.99\) and \(\pi_2=0.01\).
The fifth dimension is calculated to satisfy the five-dimensional semi-hypersphere surface equation \(X_1^2 + X_2^2 + X_3^2 + X_4^2 + X_5^2 = r^2\) where \(X_5>0\) and \(r\) is set as \(7\).
The Gaussian mixture densities could be calculated using Equation \eqref{eq:gmm} as the true density of the 5-d meta data. \autoref{fig:fivedmeta} shows a scatterplot display when animating a 5-D tour path with the R package \emph{tourr} {[}REFERENCE{]}. The round and triangular point shapes index the two mixture components \(\mathcal{N}(\pmb{\mu}_1, \pmb{\Sigma}_1)\) and \(\mathcal{N}(\pmb{\mu}_2, \pmb{\sigma}_2)\), and the colors shows the distance between the simulated \(4-D\) data point from Gaussian mixture model and the kernel means \((0, 0, 0, 0)^\prime\). The more distant from the point to the kernel cores, the lower the true densities, which shows in a darker color in \autoref{fig:fivedmeta}. It can be seen that the most distant points are in triangular shape, meaning that they are simulated from \(\mathcal{N}(\pmb{\mu}_2, \pmb{\Sigma}_2)\). The dark colors also indicate that they are the true outliers because of their low densities.

Then we initial the other 95 dimensions in the high-dimensional space as zero columns and further rotate the \(100\)-dimensional data of size \(N\) (denote the transpose of the data matrix as \(\pmb{X}_0\) with dimension \(100 \times N\)) to get rid of the zeros so that it could be passed to the manifold learning algorithms.
The rotation matrix is derived from the QR decomposition of a \(100\times 100\) matrix \(\pmb{A}\) with all components randomly generated from a uniform distribution \(\mathcal{U}(0,1)\).
For any real matrix \(\pmb{A}\) of dimension \(p\times q\), the QR decomposition could decompose the matrix into the multiplication of two matrix \(\pmb{Q}\) and \(\pmb{R}\) so that \(\pmb{A} = \pmb{QR}\), where the dimension of \(Q\) is a matrix with unit norm orthogonal vectors, \(\pmb{Q}^\prime \pmb{Q} = \pmb{I}\), and \(\pmb{R}\) is an upper triangular matrix. Matrix \(\pmb{Q}\) satisfies
\(\pmb{X}_0^\prime \pmb{X}=(\pmb{QX}_0)^\prime(\pmb{QX}_0)\), meaning that the distances between data points in \(\pmb{X}_0^\prime\) is equivalent to that of \((\pmb{QX}_0)^\prime\). Therefore, we use matrix \(Q\) as the rotation matrix for where the rotated data matrix \(\pmb{X} = (\pmb{QX}_0)^\prime\) of dimension \(N \times 100\) is now the input data for the manifold learning algorithms. Again, we set the embedding dimension to be equal to the meta data dimension \(d=5\).

@@

{\centering \includegraphics[width=0.7\linewidth]{figures/tourr_5d_semisphere}

@@

\caption{Scatterplot display of the animation of a 5-D tour path with shapes indexing the Gaussian mixture component and the colors showing the distance to the kernel cores.}\label{fig:fivedmeta}
\end{figure}
